---
title: "Getting started with predictomics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with predictomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 5, fig.height = 5
)
```

This tutorial gives a brief tour of the two main entry points in
**predictomics**: `predict_cv()`, which fits a single cross-validated
prediction pipeline, and `compare_pipelines()`, which benchmarks several
pipeline choices against a baseline and a reference model. All examples use
simulated data from `simulate_predictomics_data()`, which ships with the
package so you can regenerate (or tweak) it at any time - see `?simulate_predictomics_data`.

```{r setup}
library(predictomics)
set.seed(1)
```

## 1. Simulated data

`simulate_predictomics_data()` creates a gene expression matrix `X` organised
into genesets, a small number of which ("signal genesets") actually drive the
response `Y`; the rest are noise. It also simulates a binary `treatment` and
covariates (`age`, `sex`).

```{r}
d <- simulate_predictomics_data(n = 80, p = 200, n_genesets = 10,
                                geneset_size = 20, n_signal_genesets = 2,
                                seed = 1)

dim(d$X)
d$signal_genesets
head(d$covariates)
```

## 2. A basic `predict_cv()` pipeline

At its simplest, `predict_cv()` just needs `Y` and `X`; by default it fits an
ordinary least squares model with 10-fold cross-validation and no feature
engineering or selection.

```{r}
result <- predict_cv(Y = d$Y, X = d$X, folds = 5, verbose = FALSE)
result
```

`plot()` shows observed vs. cross-validated predicted values:

```{r}
plot(result)
```

## 3. Feature selection

With `p = 200` genes and `n = 80` samples, an unfiltered `lm` is unlikely to
generalise well. `selection_params` lets you filter to the most promising
features before fitting - here, the top 20 genes by absolute Pearson
correlation with `Y`:

```{r}
result_pearson <- predict_cv(
  Y = d$Y, X = d$X,
  selection_params = list(method = "pearson", top_n = 20),
  folds   = 5,
  verbose = FALSE
)
result_pearson
```

`plot_selection_stability()` shows how consistently each feature is selected
across the outer CV folds - a stable, informative filter should repeatedly
pick genes belonging to the true signal genesets:

```{r}
plot_selection_stability(result_pearson, plot_type = "frequency", top_n = 15)
```

## 4. Feature engineering: geneset aggregation

`engineering_params` can collapse genes into geneset-level scores before
selection/modelling - useful for interpretability and for reducing
dimensionality upfront. Here we aggregate by geneset mean, then select the
most predictive genesets:

```{r}
result_geneset <- predict_cv(
  Y = d$Y, X = d$X,
  engineering_params = list(method = "engineer", col_transform = "none",
                            genesets = d$genesets, agg_method = "mean"),
  selection_params   = list(method = "pearson", top_n = 3),
  folds   = 5,
  verbose = FALSE
)
result_geneset
```

## 5. Model choice, treatment, and covariates

`model_params$method` switches the underlying model (`"lm"`, `"glmnet"`,
`"ridge"`, `"lasso"`, `"ranger"`, `"svr"`), each tuned internally via inner
cross-validation. `treatment` and `covariates` are appended as protected
predictors, always included regardless of feature selection:

```{r}
result_glmnet <- predict_cv(
  Y = d$Y, X = d$X,
  selection_params    = list(method = "pearson", top_n = 20),
  model_params         = list(method = "glmnet", inner_folds = 5),
  treatment            = d$treatment,
  treatment_predictor  = TRUE,
  covariates           = d$covariates,
  folds   = 5,
  verbose = FALSE
)
result_glmnet

# Colour the observed-vs-predicted plot by an arbitrary variable
plot(result_glmnet, highlight = d$covariates$age, highlight_type = "continuous",
    highlight_label = "Age")
```

## 6. Baseline model

Passing `X = NULL` fits a **baseline model** using only `covariates`/
`treatment` (or, absent those, the training-fold mean of `Y`) - a built-in
null comparator for judging whether the gene expression data adds anything:

```{r}
result_baseline <- predict_cv(
  Y = d$Y, X = NULL,
  covariates = d$covariates,
  folds      = 5,
  verbose    = FALSE
)
result_baseline
```

## 7. Comparing pipelines with `compare_pipelines()`

Rather than calling `predict_cv()` repeatedly by hand, `compare_pipelines()`
fits a baseline, a reference pipeline, and a set of alternative pipelines that
vary exactly one stage (`option_type`: `"selection"`, `"engineering"`, or
`"model"`), in one call. Here we compare three filtration methods against a
Pearson-filtered `glmnet` reference:

```{r}
cmp <- compare_pipelines(
  Y = d$Y, X = d$X,
  option_type    = "selection",
  option_choices = list(
    spearman = list(method = "spearman", top_n = 20),
    variance = list(method = "variance", top_n = 20)
  ),
  reference_params = list(
    selection_params = list(method = "pearson", top_n = 20),
    model_params      = list(method = "glmnet", inner_folds = 5)
  ),
  covariates = d$covariates,
  folds      = 5,
  verbose    = FALSE
)
cmp
plot(cmp)
```

The bars show the Reference and each Alternative pipeline; the dashed line
marks the Baseline; the hatched bar marks the best-performing pipeline.

## 8. Paired pre/post designs

For repeated-measures studies (e.g. pre/post vaccination), pass a
`design = "paired"` dataset - `individual_id` and `timepoint` (0 = pre,
1 = post) unlock `engineering_params$gene_level_fc` (fold-change features),
`selection_params$dearseq_mode = "paired"` (differential expression, requires
the `dearseq` package), and `selection_params$rise_paired` (requires
`SurrogateRank`):

```{r}
d_paired <- simulate_predictomics_data(n = 40, p = 200, design = "paired", seed = 2)

result_fc <- predict_cv(
  Y = d_paired$Y, X = d_paired$X,
  engineering_params = list(method = "engineer", gene_level_fc = TRUE),
  selection_params   = list(method = "pearson", top_n = 20),
  individual_id      = d_paired$individual_id,
  timepoint          = d_paired$timepoint,
  folds   = 5,
  verbose = FALSE
)
result_fc
```

See `?predict_cv` for full details on `dearseq`- and RISE-based differential
expression selection.

## Session info

```{r}
sessionInfo()
```
