---
title: "Getting started with predictomics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with predictomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 5, fig.height = 5
)
```

This tutorial gives a brief tour of the two main entry points in
**predictomics**: `predict_cv()`, which fits a single cross-validated
prediction pipeline, and `compare_pipelines()`, which benchmarks several
pipeline choices against a baseline and a reference model. All examples use
simulated data from the internal function `simulate_predictomics_data()`.

```{r setup}
library(predictomics)
set.seed(1)
```

## 1. Simulated data

`simulate_predictomics_data()` creates a gene expression matrix `X` organised
into genesets, a small number of which ("signal genesets") actually drive the
response `Y`; the rest are noise. It also simulates a binary `treatment` and
covariates (`age`, `sex`).

```{r}
d <- simulate_predictomics_data(n = 80, p = 200, n_genesets = 10,
                                geneset_size = 20, n_signal_genesets = 2,
                                seed = 1)

dim(d$X)
d$signal_genesets
head(d$covariates)
```

## 2. A basic `predict_cv()` pipeline

At its simplest, `predict_cv()` just needs `Y` and `X`; by default it fits an
ordinary least squares model with 10-fold cross-validation and no feature
engineering or selection.

```{r}
result <- predict_cv(Y = d$Y, X = d$X, folds = 5, verbose = FALSE)
result
```

`plot()` shows observed vs. cross-validated predicted values:

```{r}
plot(result)
```

## 3. Feature selection

With `p = 200` genes and `n = 80` samples, an unfiltered `lm` is unlikely to
generalise well. `selection_params` lets you filter to the most promising
features before fitting - here, the top 20 genes by absolute Pearson
correlation with `Y`:

```{r}
result_pearson <- predict_cv(
  Y = d$Y, X = d$X,
  selection_params = list(method = "pearson", top_n = 20),
  folds   = 5,
  verbose = FALSE
)
result_pearson
```

The results are much better with the within-loop feature selection reducing the dimension prior to modelling. 

```{r}
plot(result_pearson)
```


`plot_selection_stability()` shows how consistently each feature is selected
across the outer CV folds - a stable, informative filter should repeatedly
pick genes belonging to the true signal genesets:

```{r}
plot_selection_stability(result_pearson, plot_type = "frequency", top_n = 15)
```

## 4. Feature engineering: geneset aggregation

`engineering_params` can collapse genes into geneset-level scores before
selection/modelling - useful for interpretability and for reducing
dimensionality upfront. Here we aggregate by geneset mean using the `agg_method` argument, then select the
top 3 most predictive genesets within each fold. Additionally, we apply a pre-transformation (z-score) to the gene-level features prior to this transformation using the `col_transform` argument. 

```{r}
result_geneset <- predict_cv(
  Y = d$Y, X = d$X,
  engineering_params = list(method = "engineer", col_transform = "z",
                            genesets = d$genesets, agg_method = "mean"),
  selection_params   = list(method = "pearson", top_n = 3),
  folds   = 5,
  verbose = FALSE
)
result_geneset
```
```{r}
plot(result_geneset)
```

## 5. Model choice, treatment, and covariates

`model_params$method` switches the underlying model (`"lm"`, `"glmnet"`,
`"ridge"`, `"lasso"`, `"ranger"`, `"svr"`), each tuned internally via inner
cross-validation. `treatment` and `covariates` are appended as protected
predictors, always included regardless of feature selection:

```{r}
result_glmnet <- predict_cv(
  Y = d$Y, X = d$X,
  selection_params    = list(method = "pearson", top_n = 20),
  model_params         = list(method = "glmnet", inner_folds = 5),
  treatment            = d$treatment,
  treatment_predictor  = TRUE,
  covariates           = d$covariates,
  folds   = 5,
  verbose = FALSE
)
result_glmnet
```

Let's take the opportunity to show off how we can highlight these results by covariates, controlled via the `highlight` argument in `plot()`, to examine patterns in the predictions. For a continuous covariate, we specify `highlight_type = "continuous"`. 
```{r}
# Colour the observed-vs-predicted plot by age
plot(result_glmnet, highlight = d$covariates$age, highlight_type = "continuous",
    highlight_label = "Age")
```

And for a categorical covariate, we specify `highlight_type = "categorical"`.

```{r}
plot(result_glmnet, highlight = d$covariates$sex, highlight_type = "categorical",
    highlight_label = "Sex")
```


## 6. Baseline model

Passing `X = NULL` fits a **baseline model** using only `covariates`/
`treatment` (or, absent those, the training-fold mean of `Y`) - a built-in
null comparator for judging whether the high-dimensional predictors add anything:

```{r}
result_baseline <- predict_cv(
  Y = d$Y, X = NULL,
  covariates = d$covariates,
  folds      = 5,
  verbose    = FALSE
)
result_baseline
```

## 7. Comparing pipelines with `compare_pipelines()`

Rather than calling `predict_cv()` repeatedly by hand, `compare_pipelines()`
fits a baseline pipeline, a reference pipeline, and a set of alternative pipelines that
vary exactly one stage (`option_type`: `"selection"`, `"engineering"`, or
`"model"`), in one call. Here we compare three variable selection methods against
the Pearson-correlation selection method we saw previously:

```{r}
cmp <- compare_pipelines(
  Y = d$Y, X = d$X,
  option_type    = "selection",
  option_choices = list(
    spearman = list(method = "spearman", top_n = 20),
    variance = list(method = "variance", top_n = 20)
  ),
  reference_params = list(
    selection_params = list(method = "pearson", top_n = 20),
    model_params      = list(method = "glmnet", inner_folds = 5)
  ),
  covariates = d$covariates,
  folds      = 5,
  verbose    = FALSE
)
cmp
plot(cmp, metric = "R2")
```

The bars show the Reference and each Alternative pipeline; the dashed line
marks the Baseline (i.e. covariate-only model); the hatched bar marks the best-performing pipeline.

We see that all approaches perform better than the baseline, but the variance-based feature selection performs the best, better than the reference (Pearson) and Spearman approaches. 


## Session info

```{r}
sessionInfo()
```
